---
title: "Analyses and Graphs for Biology Letters Submission - Adaptation to Common v. Rare Clones"
author: 
date: "5/27/2020"
output: html_document
---

Packages
```{r}
library(ggplot2)
library(ggthemes)
library(lme4)
library(reshape2)
library(afex)
library(Hmisc)
```

fit plot function
```{r}
fitplot<-function(x){
  plot(fitted(x), residuals(x), xlab = "Fitted Values", ylab = "Residuals")
abline(h = 0, lty = 2)
lines(smooth.spline(fitted(x), residuals(x)))
}
```

Data set - ancestral values
```{r}
datAnc<-read.csv("CommonClone_ancestor.csv",header=T)
```

Data set - adaptation to homogeneous host populations
```{r}
datHom<-read.csv("CommonClone_homogeneous.csv",header=T)
```

Data set - adaptation in heterogeneous host populations; common clone analysis
```{r}
dat<-read.csv("CommonClone_dataset.csv",header=T)
dat$Line=factor(dat$Line)
dat$Rep=factor(dat$Rep)
summary(dat)
dat<- within(dat, Frequency <- relevel(Frequency, ref = "common"))
```

Figure S1: ancestral values; electronic supplementary material
```{r}
# summarize
ancM<-tapply(datAnc$Mortality,list(datAnc$Host),mean)
ancSE<-tapply(datAnc$Mortality,list(datAnc$Host),sd)/sqrt(tapply(datAnc$Mortality,list(datAnc$Host),length))
anc<-cbind(ancM,ancSE)
host<-c("CB4856","ewIR 68","LTM1","N2")
anc<-as.data.frame(anc)

theme_set(theme_bw())
FigS1<-ggplot(data=anc, aes(x=host, y=ancM)) +
    geom_errorbar(aes(ymin=ancM-ancSE, ymax=ancM+ancSE), width=0,size=1,
                  position=position_dodge(width=0.25)) +
    geom_point(size=10, shape=21,color="#000000",stroke=2,position=position_dodge(width=0.25),fill="gray60") +
    xlab("") + ylab("Mean Mortality Rate") +
    ylim(0.5, 1) +
  theme(axis.text.x = element_text(color="black", 
                           size=14),
          axis.text.y = element_text(color="black", 
                           size=24),
        axis.title.y=element_text(face="bold", color="black", 
                           size=20),
        legend.text = element_text(size=14),
        legend.title = element_text(size=20))+
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())

# statistical test of ancestral differences
overdisp_fun <- function(model) {
  rdf <- df.residual(model)
  rp <- residuals(model,type="pearson")
  Pearson.chisq <- sum(rp^2)
  prat <- Pearson.chisq/rdf
  pval <- pchisq(Pearson.chisq, df=rdf, lower.tail=FALSE)
  c(chisq=Pearson.chisq,ratio=prat,rdf=rdf,p=pval)
}

# supplementary analyses
datAnc$Host = relevel(datAnc$Host, ref="ewIR 68")
ancM1<-glmer(cbind(Dead,Live)~Host + (1|Assay),family="binomial",data=datAnc)
overdisp_fun(ancM1) # yes
datAnc$ID=factor(datAnc$ID)
ancM2<-glmer(cbind(Dead,Live)~Host + (1|Assay) + (1|ID),family="binomial",data=datAnc) # overfit
ancM3<-glmer(cbind(Dead,Live)~Host + (1|ID),family="binomial",data=datAnc)
ancM4<-update(ancM3,.~.-Host)
anova(ancM3,ancM4)
summary(ancM3)
```

Figure S2: adaptation to homogeneous host populations
Table S1
```{r}
mean<-tapply(datHom$Mortality,list(datHom$Host,datHom$TreatmentGroup),mean) # Table S1
se<-tapply(datHom$Mortality,list(datHom$Host,datHom$TreatmentGroup),sd)/sqrt(tapply(datHom$Mortality,list(datHom$Host,datHom$TreatmentGroup),length))
diff1=(mean[,3]-mean[,2])
diff=diff1/mean[,2] # proportional increase from control
diffse1=sqrt(se[,2]^2+se[,3]^2)
diffse=diff*sqrt(((diffse1/diff1)^2)+((se[,2]/mean[,2])^2))

Hom<-cbind(diff,diffse)
host<-c("CB4856","ewIR 68","LTM1","N2")
Hom<-as.data.frame(Hom)

FigS2<-ggplot(data=Hom, aes(x=host, y=diff)) +
      geom_hline(yintercept=0.0,linetype="dashed",color="gray40",size=1.4)+
    geom_errorbar(aes(ymin=diff-diffse, ymax=diff+diffse), width=0,size=1,
                  position=position_dodge(width=0.25)) +
    geom_point(size=8, shape=21,color="#000000",stroke=2,position=position_dodge(width=0.25),fill="gray60") +
    xlab("") + ylab("Mean Mortality Rate") +
    ylim(-0.05, 0.8) +
  theme(axis.text.x = element_text(color="black", 
                           size=14),
          axis.text.y = element_text(color="black", 
                           size=24),
        axis.title.y=element_text(face="bold", color="black", 
                           size=20),
        legend.text = element_text(size=14),
        legend.title = element_text(size=20))+
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())
```

data manipulation
```{r}
dat$Line=factor(dat$Line)
dat$Rep=factor(dat$Rep)
summary(dat)
dat<- within(dat, Frequency <- relevel(Frequency, ref = "common"))

# calculating change from ancestral values
# mortality rates according to rare/common by each host genotype
# proportional change
change<-tapply(dat$PropChange,list(dat$Line,dat$Frequency),mean,na.rm=T)
change<-data.frame(change)
change[,3]=seq(1,36)
colnames(change)=c("common","rare","id")
#melt dataframe
changedf<- melt(change, id.vars="id",variable.name="Frequency", value.name="change")
changedf<- within(changedf, Frequency <- relevel(Frequency, ref = "rare"))
changedf$Host=c(rep("N2",12),rep("LTM1",12),rep("CB4856",6),rep("ewir68",6),rep("LTM1",12),rep("N2",12),rep("ewir68",6), rep("CB4856",6))
changedf$Common=c(rep("N2",12),rep("LTM1",12),rep("CB4856",6),rep("ewir68",6))

# summary by common host
changemean<-tapply(changedf$change,list(changedf$Common,changedf$Frequency),mean)
changemean<-data.frame(changemean)
changese<-tapply(changedf$change,list(changedf$Common,changedf$Frequency),sd)/sqrt(tapply(changedf$change,list(changedf$Common,changedf$Frequency),length))
changemean$Common=c("CB4856","ewir68","LTM1","N2")

# provides proportion change and SE by line - referenced in main text lines 165-171
sumP<-melt(changemean, id.vars="Common")
sumP$se<-c(changese[1:4,1],changese[1:4,2])

```

Figure 2
```{r}
# all lines, by panel
Fig2 <-ggplot()+ 
    geom_hline(yintercept=0.0,linetype="dashed",color="gray40",size=1.4)+
    geom_line(data=changedf,aes(x=Frequency, y=change, group=id),size=1.05,color="gray60")+
    geom_point(data=changedf,aes(x=Frequency, y=change, group=id),size=4, shape=21,color="gray60",fill="gray90",stroke=1.5) + 
    xlab("") + ylab("") +
  scale_y_continuous(breaks=seq(-0.4,0.2,0.1),limits=c(-0.42,0.19))+
      geom_errorbar(dat=sumP,aes(x=variable,ymin=value-se, ymax=value+se), width=0,size=1)+  
  geom_line(data=sumP,aes(x=variable,y=value,group=Common),size=1.1,color="black")+
  geom_point(data=sumP,aes(x=variable,y=value,group=Common),size=6,shape=21,color="gray10",fill=c("#b2182b","#fddbc7","#d1e5f0","#2166ac","#fddbc7","#b2182b","#2166ac","#d1e5f0"),stroke=2)+
  facet_grid(~Common)+
  theme(axis.text.x = element_text(color="black", 
                           size=14),
          axis.text.y = element_text(color="black", 
                           size=24),
        legend.text = element_text(size=14),
        legend.title = element_text(size=20))+
    theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())
Fig2
```

Change from ancestor - linear model; Table S2
```{r}
dat$Frequency <- factor(dat$Frequency,c("rare","common") )
dat$AssayHost <-factor(dat$AssayHost,c("ewIR68","CF3-wt-inv","N2","CB4856")) 

# random effects: parasite line, nested within treatment, nested within experiment; assay number
# predictors: Assayed host genotype (LTM1, N2, CB4856); host genotype frequency during exp evolution (rare or common)
lmfit<-lmer(DiffAnc~AssayHost*Frequency  + (1|Experiment/ExperimentalTreatment/Line) + (1|AssayBlock), data=dat)
# can't estimate parameters - remove experiment as random effect
lmfit2<-lmer(DiffAnc~AssayHost * Frequency +  (1|ExperimentalTreatment/Line)+(1|AssayBlock), data=dat,control=lmerControl(optimizer="bobyqa",optCtrl=list(maxfun=2e5)))
anova(lmfit2) # Table S3A
summary(lmfit2)
# remove random effects
# remove assayblock
lmfit3<-lmer(DiffAnc~AssayHost * Frequency +  (1|ExperimentalTreatment/Line), data=dat, control=lmerControl(optimizer="bobyqa",optCtrl=list(maxfun=2e5)))
anova(lmfit2,lmfit3) 
# remove experimental treatment
lmfit4<-lmer(DiffAnc~AssayHost * Frequency +  (1|Line)+(1|AssayBlock), data=dat,control=lmerControl(optimizer="bobyqa",optCtrl=list(maxfun=2e5)))
anova(lmfit2,lmfit4)
anova(lmfit4) #Table S3D
# both
lmfit5<-lmer(DiffAnc~AssayHost * Frequency +  (1|Line), data=dat,control=lmerControl(optimizer="bobyqa",optCtrl=list(maxfun=2e5))) # given individual IDs of lines, this is effectively the same as nesting
anova(lmfit4,lmfit5)

# remove interaction
dat$AssayHost=relevel(dat$AssayHost, ref="CB4856")
dat$Frequency=relevel(dat$Frequency, ref="common")
lmfit6<-lmer(DiffAnc~AssayHost + Frequency +  (1|ExperimentalTreatment/Line)+(1|AssayBlock), data=dat, control=lmerControl(optimizer="bobyqa",optCtrl=list(maxfun=2e5))) # retaining treatment because of relevance to structure
fitplot(lmfit6)
anova(lmfit6) #Table S3B
summary(lmfit6) # Table S3C
anova(lmfit2,lmfit6)
```

alternative analysis 1
non-parametric approach;Figure 2 and Table S2
```{r}
# calculating change from ancestral values
# mortality rates according to rare/common by each host genotype
# difference from the ancestor
change2<-tapply(dat$DiffAnc,list(dat$Line,dat$Frequency),mean,na.rm=T)
change2<-data.frame(change2)
change2[,3]=seq(1,36)
colnames(change2)=c("common","rare","id")
#melt dataframe
changedf2<- melt(change2, id.vars="id",variable.name="Frequency", value.name="change")
changedf2<- within(changedf2, Frequency <- relevel(Frequency, ref = "rare"))
changedf2$Host=c(rep("N2",12),rep("LTM1",12),rep("CB4856",6),rep("ewir68",6),rep("LTM1",12),rep("N2",12),rep("ewir68",6), rep("CB4856",6))
changedf2$Common=c(rep("N2",12),rep("LTM1",12),rep("CB4856",6),rep("ewir68",6))
# is diffanc shifted from 0?
# all 36 lines
wilcox.test(change2$rare,change2$common,paired=T,alternative="two.sided")
wilcox.test(change2$rare-change2$common,alternative="two.sided")

# individual tests
# rare hosts
changedf2_rare<-subset(changedf2,changedf$Frequency=="rare")

# ewir68 rare
ewIR68_rare<-subset(changedf2_rare,changedf2_rare$Host=="ewir68")
wilcox.test(ewIR68_rare$change,mu=0,alternative="two.sided")
# CB4856 rare
CB4856_rare<-subset(changedf2_rare,changedf2_rare$Host=="CB4856")
wilcox.test(CB4856_rare$change)
# N2 rare
N2_rare<-subset(changedf2_rare,changedf2_rare$Host=="N2")
wilcox.test(N2_rare$change)
# LTM1 rare
LTM1_rare<-subset(changedf2_rare,changedf2_rare$Host=="LTM1")
wilcox.test(LTM1_rare$change)

# common hosts
changedf2_common<-subset(changedf2,changedf$Frequency=="common")

# ewir68 common
ewIR68_common<-subset(changedf2_common,changedf2_common$Host=="ewir68")
wilcox.test(ewIR68_common$change)
# CB4856 common
CB4856_common<-subset(changedf2_common,changedf2_common$Host=="CB4856")
wilcox.test(CB4856_common$change)
# N2 common
N2_common<-subset(changedf2_common,changedf2_common$Host=="N2")
wilcox.test(N2_common$change)
# LTM1 common
LTM1_common<-subset(changedf2_common,changedf2_common$Host=="LTM1")
wilcox.test(LTM1_common$change)

#calculating W+,W- 
diff <- c(change2$rare-change2$common) 
diff <- diff[ diff!=0 ] 
diff.rank <- rank(abs(diff)) 
diff.rank.sign <- diff.rank * sign(diff) 
ranks.pos <- sum(diff.rank.sign[diff.rank.sign > 0]) 
ranks.neg <- -sum(diff.rank.sign[diff.rank.sign < 0])
ranks.pos 
ranks.neg
```

alternative analysis 2
meta-analysis
Figure S3
```{r}
datMA<-read.csv("CommonClone_meta.csv",header=T)
datMA<-datMA[1:4,8:21]

library(metafor)
# overall relationship
res<-rma(yi = g, vi=Vg, data = datMA)

# Figure S3
forest(res,cex=2)
```

summary of proportional change
```{r}
# estimates of effect
mean(change$common)
mean(change$rare)
sd(change$common)/sqrt(length(change$common))
sd(change$rare)/sqrt(length(change$rare))

mean(change$common-change$rare)
median(change$common-change$rare)
diff<-change$common-change$rare
quartile <- as.numeric(cut2(diff, g=4))   #as.numeric to number the factors
Bounds <- cut2(diff, g=4)
```
